---
title: "Analysis Title"
author: "smjb"
date: "July 13, 2015"
output: html_document
---
#preamble - TODO
    
    You may have other sections in your analysis, but Data Processing and Results are required.


##Megatitle - TODO

Part of Reproductible Research - Peer Assignment 2

##Synopsis - TODO
! 10 sentence max

##Analysis objective

This analysis is made to address the following questions:

1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

2. Across the United States, which types of events have the greatest economic consequences?

This report is intended for a government or municipal manager who might be responsible for preparing for severe weather events and will need to prioritize resources for different types of events. However, this document  does not make any specific recommendations.

##Data Processing - TODO

```{r cache=TRUE, message=FALSE}

# loading libraries

library(dplyr)
library(lattice)
library(lubridate)

## basic configuration

nb_top_event = 20

bzstormfile <- "stormdata.bz2" # just for file name consistency

if(!file.exists(bzstormfile)) { # no need to download everytime we run
   download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", "bzstormfile")
}

#read.csv can read bzfiles directly. Hooray!

storm <- read.csv(bzstormfile)

dstorm <- tbl_df(storm)
warning("BGN_TIME and END_TIME format are not consistent across observation. Needs cleanup if using it")
# to convert ABCD format into AB:CD format, use the command below
# xtime = gsub('^([0-9]{2})([0-9]+)$', '\\1:\\2',xtime)
# dstorm <- mutate(dstorm, start_date = mdy_hms(BGN_DATE), stop_date = mdy_hms(END_DATE), ....)
```

## Data contained in file
```{r cache=TRUE}
str(storm)

```
#### fields indicating harm to health 

1. FATALITIES
2. INJURIES

#### fields indicating economic consequences

1. PROPDMG
2. PROPDMGEXP
3. CROPDMG
4. CROPDMGEXP

```{r cache=TRUE, fig.width=800, fig.height=800}
#freeing up some memory by selecting necessary columns only

storm_damage <- dstorm %>% select(STATE, EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP) %>% group_by(EVTYPE) 

# stat_human_injury <- storm_damage %>% 
#                         mutate(total_fatal_event = sum(FATALITIES>0), total_fatalities = sum(FATALITIES),
#                                 total_harmful_event = sum(INJURIES>0), total_injuries = sum(INJURIES)) %>% 
#                         arrange(desc(total_fatalities), desc(total_injuries))


fatal_event <- storm_damage %>% summarise(total_fatalities = sum(FATALITIES), total_fatal_event=sum(FATALITIES>0)) %>% arrange(desc(total_fatalities), desc(total_fatal_event))

harmful_event <- storm_damage %>% summarise(total_fatalities = sum(FATALITIES), total_fatal_event=sum(FATALITIES>0),
                                            total_injuries = sum(INJURIES), total_injury_event=sum(INJURIES>0), 
                                            fatalities_per_event = total_fatalities/total_fatal_event, 
                                            injuries_per_event = total_injuries/total_injury_event) %>% 
    arrange(desc(total_fatalities), desc(total_injuries), desc(total_fatal_event), desc(total_injury_event))

top_harmful_event <-  harmful_event[1:nb_top_event,]

x_max = max(top_harmful_event$total_fatalities);

par(mar=c(4,14,8,4), cex=0.75, las=1, mfrow=c(2,2))
barplot(top_harmful_event$total_fatalities, 
        horiz=TRUE, 
        names.arg = top_harmful_event$EVTYPE, 
        xlim = c(0, ceiling(max(top_harmful_event$total_fatalities)/1000)*1000), 
        ylim=c(0,nb_top_event), 
        main=paste0("Total Fatalities of the top ", nb_top_event, " Event Type"), 
        xlab ="Total Fatalities", 
        ylab = ""
    )

x_max = max(top_harmful_event$total_injuries);

barplot(top_harmful_event$total_injuries, 
        horiz=TRUE, 
        names.arg = top_harmful_event$EVTYPE, 
        xlim = c(0, ceiling(x_max/10^floor(log10(x_max)))*10^floor(log10(x_max))), 
        
        ylim=c(0,nb_top_event), 
        main=paste0("Total Injuries of the top ", nb_top_event, " Event Type"), 
        xlab ="Total Fatalities", 
        ylab = ""
    )


#verify that there is no missing values in the selected subset

number.of.na <- length(which(is.na(harmful_event)))
number.of.na

```
## Results 
### Harm to population

```{r cache=TRUE}



```


    The analysis document must have at least one figure containing a plot.

    Your analyis must have no more than three figures. Figures may have multiple plots in them (i.e. panel plots), but there cannot be more than three figures total.


